#Impact of Storm Events in the United States

### Synopsis

This is an analysis of storm event data from the National Oceanic and Atmospheric Administration. The database can be found at [This location](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and contains data on events from 1950 to 2010. Since the early years of observations are not as complete, for the purposes of this analysis, we will (arbitrarily) only use observations from 1990 to 2010. 

This analysis will try to answer the question of which type of event is most harmful to society. The answer will be broken into two approaches. 

1. Which event type is most harmful to the public health?

2. Which event type causes the most economic damage?

### Data Preparation
If necessary, we can use download.file to get the compressed file into our working directory.
```{r, eval=F}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "stormdata.csv.bz2")
```

Read.csv() is capable of extracting bz2 files natively, so after downloading, we can read the storm data directly into our workspace.

```{r, eval=T, cache=T}
data <- read.csv("stormdata.csv.bz2")
```

```{r, echo=FALSE, eval=FALSE}
data <- read.csv("repdata-data-StormData.csv")
```

Let's take a quick look at the structure of our data.

```{r}
str(data)
```

We have over 900,000 observations of 37 variables. Some of these variables are specific to the storm events, like the f ratings of tornadoes. A good starting point would be removing the variables that only apply to specific events, and to filter for dates that are after January 1 1990. I will use dplyr where possible for my data transformations.

```{r, cache=TRUE}
library(dplyr)

data <- data %>% mutate(BGN_DATE= as.Date(BGN_DATE, format= "%m/%d/%Y")) %>% 
        filter(BGN_DATE >= 1/1/1990) %>% #Keep only events after the given date.
        select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

head(data)
```


In order to get proper amounts for the Property and Crop damage, we need to multiply the number in the value columns by the appropriate values indicated by the "EXP" columns. However, some entries don't match the codebook, which allows for an H, K, M or B to know what the correct multiplier is that should be applied to the DMG columns.

```{r, cache=TRUE}
#How many of each invalid EXPs are present?
data %>% group_by(PROPDMGEXP) %>% summarize(count=n())
data %>% group_by(CROPDMGEXP) %>% summarize(count=n())
```

I decided that removing the ambiguous entries would not have a significant impact on the analysis, since they are relatively few.

I'll write a simple function to transform the CROPDMG and PROPDMG variables and apply it with mutate():
```{r, cache=TRUE}
cleanexps <- function(DMG,EXP){
        EXP <- toupper(EXP)
        multiplier<- c(H=100,K=1000,M=1000000,B=1000000000)
        for(i in seq_along(EXP)){
                if(!(EXP[i] %in% names(multiplier))){
                        DMG[i] <- (0*DMG[i])
                        }
                else{
                        DMG[i] <- as.numeric(multiplier[EXP[i]] * DMG[i])
                }
        }
        return(DMG)
}
data <- data %>% mutate(CROPDMG = cleanexps(CROPDMG,CROPDMGEXP), PROPDMG= cleanexps(PROPDMG,PROPDMGEXP)) %>% select(-CROPDMGEXP, -PROPDMGEXP)
```

Let's check the results.
```{r}
summary(data$CROPDMG)

summary(data$PROPDMG)

```

There are also issues with the event type variable critical to our analysis.

```{r}
head(unique(data$EVTYPE),100)
```

This is only the first 100 unique entries, but displays several of the problems with this variable. For example, hurricanes are named individually instead of as a group. Also, some lines include several event types of differing tiers. For example, the proximate cause of damage when tornadoes coincide with hail could be assumed that the tornado. When hail coincides with wind, however you could assume the hail is doing the most damage. With that in mind, a function will help us sort out the variable.

```{r}
cleantype <- function(EVTYPE){
        
}

```